---
title: "260508_Stoats_DEG"
author: "Abby Bradley"
date: "2026-05-14"
output:
  html_document: default
  pdf_document: default
---

## step 1: load libraries
```{r}
library(DESeq2)
library(dplyr)
library(tidyverse)
library(biomaRt)
```



## step 2: read in files
row.names = 1: col1 has the rownames
note: counts themselves are technically filtered

### counts themselves

```{r}
data_raw <- read.csv(file = "C:/Users/abbyb/UCLA/CASB_185/GSE213001_filtered_counts.txt",
                     row.names = 1,
                     header = TRUE)
head(data_raw)
```


### metadata
```{r}
metadata <- read.csv(file = "/Users/abbyb/UCLA/CASB_185/GSE213001_filtered_metadata.txt",
                     row.names = 1,
                     header = TRUE)
head(metadata)
```


brief pre filtering, filtering out low-count genes
```{r}
# total gene counts >100
data <- data_raw[rowSums(data_raw[,c(1:103)]) >= 100, ]

print("raw gene number:")
dim(data_raw)

print("number of genes with sum(counts) >100 across all samples:")
dim(data)
```

### remove rows with NA in grouping variable
```{r}
# metadata <- metadata[!is.na(metadata$sample_group), ]
# 
# # Keep only the columns in 'data' that correspond to the rows in metadata
# data <- data[, rownames(metadata)]
```

# step 3: DEGs

# NDC apex vs. IPF apex
## setup
### new metadata/data
first, make new metadata with only samples from NDC apex or IPF apex
```{r}
print("Number of samples total:")
dim(metadata)

# NAvIA_list grabs the sample names with the thing we want to do our DEG on
NAvIA_list <- rownames(metadata[metadata$group == "IPF.Apex" | metadata$group == "NDC.Apex", ])

NAvIA_metadata <- metadata[NAvIA_list,]

# Drop unused factor levels
NAvIA_metadata$group <- factor(NAvIA_metadata$group)

print("Number of samples for NAvIA DEG:")
dim(NAvIA_metadata)

# check it worked
head(NAvIA_metadata)

# Verify the levels now
levels(NAvIA_metadata$group)
```

then, subset data by that metadata
```{r}
NAvIA_data <- data[, NAvIA_list]

if (all(colnames(NAvIA_data) == rownames(NAvIA_metadata))) {
  print("Yes!") 
} else {
  print("No :(")
}

head(NAvIA_data)
```

### run DESeq from dataset
```{r}
# create DESeq data set object
NAvIA_dds <- DESeqDataSetFromMatrix(countData = NAvIA_data,
                              colData = NAvIA_metadata,
                              design = ~ group)

# look data object characteristics
NAvIA_dds
```

now relevel w/ NDC apex as the ctrl
```{r}
NAvIA_dds$group <- relevel(NAvIA_dds$group, ref = 'NDC.Apex')
levels(NAvIA_dds$group)
```
## DEG
now DESeq!
```{r}
NAvIA_dds <- DESeq(NAvIA_dds)
NAvIA_dds
```

### extract DEG results
pval = 0.05
ordered smallest to greatest p adjusted
```{r}
# grab results
NAvIA_dds <- results(NAvIA_dds, contrast=c("group","IPF.Apex","NDC.Apex"), alpha=0.05)

# order + view results
NAvIA_dds.ord <- NAvIA_dds[order(NAvIA_dds$padj),]
head(NAvIA_dds.ord)
```
## change names
### change ensemble IDs to gene names!
first, make NAvIA_dds.ord into a df
```{r}
NAvIA_DEG <- data.frame(NAvIA_dds.ord)
```

now do whatever evil magic is going on here

```{r}
# Extract the column with geneid
gene_id <-  rownames(NAvIA_DEG)

# Get the gene symbols for all gene_ids
ensembl <- useMart("ensembl")
ensembl_dataset <- useDataset("hsapiens_gene_ensembl", mart = ensembl)
gene_names <- getBM(attributes = c("ensembl_gene_id", "external_gene_name", "gene_biotype"),
          filters = "ensembl_gene_id",
          values = gene_id,
          mart = ensembl_dataset)
# View(gene_names)
```

```{r}
NAvIA_DEG_final <- NAvIA_DEG %>%
  tibble::rownames_to_column(var = "ensembl_gene_id") %>%
  inner_join(gene_names, by = "ensembl_gene_id") %>%
  filter(!is.na(external_gene_name) & external_gene_name != "") %>%
  column_to_rownames(var = "external_gene_name") %>%
  # Use base R indexing here instead of select()
  .[, !names(.) %in% c("ensembl_gene_id")]

# remove NAs
NAvIA_DEG_final <- na.omit(NAvIA_DEG_final)

# verify it worked
dim(NAvIA_DEG) #og
dim(NAvIA_DEG_final) #after filtering
# check rownames are now gene names
head(rownames(NAvIA_DEG_final))
```

## save results
```{r}
# write.csv(NAvIA_DEG_final,
#           file = "/Users/abbyb/UCLA/CASB_185/260513_Stoats_NAvIA_DEG_Fixed.csv")
```


# NDC base vs. IPF base
## setup
### new metadata/data
first, make new metadata with only samples from NDC base or IPF base
```{r}
print("Number of samples total:")
dim(metadata)

# NBvIB_list grabs the sample names with the thing we want to do our DEG on
NBvIB_list <- rownames(metadata[metadata$group == "NDC.Base" | metadata$group == "IPF.Base", ])

NBvIB_metadata <- metadata[NBvIB_list,]

# Drop unused factor levels
NBvIB_metadata$group <- factor(NBvIB_metadata$group)

print("Number of samples for NBvIB DEG:")
dim(NBvIB_metadata)

# check it worked
head(NBvIB_metadata)

# Verify the levels now
levels(NBvIB_metadata$group)
```


then, subset data by that metadata
```{r}
NBvIB_data <- data[, NBvIB_list]
if (all(colnames(NBvIB_data) == rownames(NBvIB_metadata))) {
  print("Yes!") 
} else {
  print("No :(")
}

head(NBvIB_data)
```

### run DESeq from dataset
```{r}
# create DESeq data set object
NBvIB_dds <- DESeqDataSetFromMatrix(countData = NBvIB_data,
                              colData = NBvIB_metadata,
                              design = ~ group)

# look data object characteristics
NBvIB_dds
```

now relevel w/ NDC base as the ctrl
```{r}
NBvIB_dds$group <- relevel(NBvIB_dds$group, ref = 'NDC.Base')
levels(NBvIB_dds$group)
```
## DEG
now DESeq!
```{r}
NBvIB_dds <- DESeq(NBvIB_dds)
NBvIB_dds
```

### extract DEG results
pval = 0.05
ordered smallest to greatest p adjusted
```{r}
# grab results
NBvIB_dds <- results(NBvIB_dds, contrast=c("group","IPF.Base","NDC.Base"), alpha=0.05)

# order + view results
NBvIB_dds.ord <- NBvIB_dds[order(NBvIB_dds$padj),]
head(NBvIB_dds.ord)
```

## change names
### change ensemble IDs to gene names!
first, make NBvIB_dds.ord into a df
```{r}
NBvIB_DEG <- data.frame(NBvIB_dds.ord)
```

now do whatever evil magic is going on here

```{r}
# Extract the column with geneid
gene_id <-  rownames(NBvIB_DEG)

# Get the gene symbols for all gene_ids
ensembl <- useMart("ensembl")
ensembl_dataset <- useDataset("hsapiens_gene_ensembl", mart = ensembl)
gene_names <- getBM(attributes = c("ensembl_gene_id", "external_gene_name", "gene_biotype"),
          filters = "ensembl_gene_id",
          values = gene_id,
          mart = ensembl_dataset)
# View(gene_names)
```

```{r}
NBvIB_DEG_final <- NBvIB_DEG %>%
  tibble::rownames_to_column(var = "ensembl_gene_id") %>%
  inner_join(gene_names, by = "ensembl_gene_id") %>%
  filter(!is.na(external_gene_name) & external_gene_name != "") %>%
  column_to_rownames(var = "external_gene_name") %>%
  # Use base R indexing here instead of select()
  .[, !names(.) %in% c("ensembl_gene_id")]

# remove NAs
NBvIB_DEG_final <- na.omit(NBvIB_DEG_final)

# verify it worked
dim(NBvIB_DEG) #og
dim(NBvIB_DEG_final) #after filtering
# check rownames are now gene names
head(rownames(NBvIB_DEG_final))
```

## save results
```{r}
# write.csv(NBvIB_DEG_final,
#           file = "/Users/abbyb/UCLA/CASB_185/260513_Stoats_NBvIB_DEG_Fixed.csv")
```



# IPF apex vs. IPF base
## setup
### new metadata/data
first, make new metadata with only samples from IPF apex or IPF base
```{r}
print("Number of samples total:")
dim(metadata)

# IAvIB_list grabs the sample names with the thing we want to do our DEG on
IAvIB_list <- rownames(metadata[metadata$group == "IPF.Apex" | metadata$group == "IPF.Base", ])

IAvIB_metadata <- metadata[IAvIB_list,]

# Drop unused factor levels
IAvIB_metadata$group <- factor(IAvIB_metadata$group)

print("Number of samples for IAvIB DEG:")
dim(IAvIB_metadata)

# check it worked
head(IAvIB_metadata)

# Verify the levels now
levels(IAvIB_metadata$group)
```


then, subset data by that metadata
```{r}
IAvIB_data <- data[, IAvIB_list]
if (all(colnames(IAvIB_data) == rownames(IAvIB_metadata))) {
  print("Yes!") 
} else {
  print("No :(")
}

head(IAvIB_data)
```

### run DESeq from dataset
```{r}
# create DESeq data set object
IAvIB_dds <- DESeqDataSetFromMatrix(countData = IAvIB_data,
                              colData = IAvIB_metadata,
                              design = ~ group)

# look data object characteristics
IAvIB_dds
```

now relevel w/ IPF apex as the ctrl
```{r}
IAvIB_dds$group <- relevel(IAvIB_dds$group, ref = 'IPF.Apex')
levels(IAvIB_dds$group)
```

## DEG
now DESeq!
```{r}
IAvIB_dds <- DESeq(IAvIB_dds)
IAvIB_dds
```

### extract DEG results
pval = 0.05
ordered smallest to greatest p adjusted
```{r}
# grab results
IAvIB_dds <- results(IAvIB_dds, contrast=c("group","IPF.Base","IPF.Apex"), alpha=0.05)

# order + view results
IAvIB_dds.ord <- IAvIB_dds[order(IAvIB_dds$padj),]
head(IAvIB_dds.ord)
```
## change names
### change ensemble IDs to gene names!
first, make IAvIB_dds.ord into a df
```{r}
IAvIB_DEG <- data.frame(IAvIB_dds.ord)
```

now do whatever evil magic is going on here

```{r}
# Extract the column with geneid
gene_id <-  rownames(IAvIB_DEG)

# Get the gene symbols for all gene_ids
ensembl <- useMart("ensembl")
ensembl_dataset <- useDataset("hsapiens_gene_ensembl", mart = ensembl)
gene_names <- getBM(attributes = c("ensembl_gene_id", "external_gene_name", "gene_biotype"),
          filters = "ensembl_gene_id",
          values = gene_id,
          mart = ensembl_dataset)
# View(gene_names)
```

```{r}
IAvIB_DEG_final <- IAvIB_DEG %>%
  tibble::rownames_to_column(var = "ensembl_gene_id") %>%
  inner_join(gene_names, by = "ensembl_gene_id") %>%
  filter(!is.na(external_gene_name) & external_gene_name != "") %>%
  column_to_rownames(var = "external_gene_name") %>%
  # Use base R indexing here instead of select()
  .[, !names(.) %in% c("ensembl_gene_id")]

# remove NAs
IAvIB_DEG_final <- na.omit(IAvIB_DEG_final)

# verify it worked
dim(IAvIB_DEG) #og
dim(IAvIB_DEG_final) #after filtering
# check rownames are now gene names
head(rownames(IAvIB_DEG_final))
```

## save results
```{r}
# write.csv(IAvIB_DEG_final,
#           file = "/Users/abbyb/UCLA/CASB_185/260513_Stoats_IAvIB_DEG_Fixed.csv")
```



# NDC apex vs. NDC base
## setup
### new metadata/data
first, make new metadata with only samples from NDc apex or NDc base
```{r}
print("Number of samples total:")
dim(metadata)

# NAvNB_list grabs the sample names with the thing we want to do our DEG on
NAvNB_list <- rownames(metadata[metadata$group == "NDC.Apex" | metadata$group == "NDC.Base", ])

NAvNB_metadata <- metadata[NAvNB_list,]

# Drop unused factor levels
NAvNB_metadata$group <- factor(NAvNB_metadata$group)

print("Number of samples for NAvNB DEG:")
dim(NAvNB_metadata)

# check it worked
head(NAvNB_metadata)

# Verify the levels now
levels(NAvNB_metadata$group)
```


then, subset data by that metadata
```{r}
NAvNB_data <- data[, NAvNB_list]
if (all(colnames(NAvNB_data) == rownames(NAvNB_metadata))) {
  print("Yes!") 
} else {
  print("No :(")
}

head(NAvNB_data)
```

### run DESeq from dataset
```{r}
# create DESeq data set object
NAvNB_dds <- DESeqDataSetFromMatrix(countData = NAvNB_data,
                              colData = NAvNB_metadata,
                              design = ~ group)

# look data object characteristics
NAvNB_dds
```

now relevel w/ IPF apex as the ctrl
```{r}
NAvNB_dds$group <- relevel(NAvNB_dds$group, ref = 'NDC.Apex')
levels(NAvNB_dds$group)
```

## DEG
now DESeq!
```{r}
NAvNB_dds <- DESeq(NAvNB_dds)
NAvNB_dds
```

### extract DEG results
pval = 0.05
ordered smallest to greatest p adjusted
```{r}
# grab results
NAvNB_dds <- results(NAvNB_dds, contrast=c("group","NDC.Base","NDC.Apex"), alpha=0.05)

# order + view results
NAvNB_dds.ord <- NAvNB_dds[order(NAvNB_dds$padj),]
head(NAvNB_dds.ord)
```
## change names
### change ensemble IDs to gene names!
first, make NAvNB_dds.ord into a df
```{r}
NAvNB_DEG <- data.frame(NAvNB_dds.ord)
```

now do whatever evil magic is going on here

```{r}
# Extract the column with geneid
gene_id <-  rownames(NAvNB_DEG)

# Get the gene symbols for all gene_ids
ensembl <- useMart("ensembl")
ensembl_dataset <- useDataset("hsapiens_gene_ensembl", mart = ensembl)
gene_names <- getBM(attributes = c("ensembl_gene_id", "external_gene_name", "gene_biotype"),
          filters = "ensembl_gene_id",
          values = gene_id,
          mart = ensembl_dataset)
# View(gene_names)
```

```{r}
NAvNB_DEG_final <- NAvNB_DEG %>%
  tibble::rownames_to_column(var = "ensembl_gene_id") %>%
  inner_join(gene_names, by = "ensembl_gene_id") %>%
  filter(!is.na(external_gene_name) & external_gene_name != "") %>%
  column_to_rownames(var = "external_gene_name") %>%
  # Use base R indexing here instead of select()
  .[, !names(.) %in% c("ensembl_gene_id")]

# remove NAs
NAvNB_DEG_final <- na.omit(NAvNB_DEG_final)

# verify it worked
dim(NAvNB_DEG) #og
dim(NAvNB_DEG_final) #after filtering
# check rownames are now gene names
head(rownames(NAvNB_DEG_final))
```

## save results
```{r}
# write.csv(NAvNB_DEG_final,
#           file = "/Users/abbyb/UCLA/CASB_185/260513_Stoats_NAvNB_DEG_Fixed.csv")
```



